---
title: "Choropleth Mapping with R"
author: "Guofang Ji"
date: "March 6, 2026"
date-modified: "`r Sys.Date()`"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
editor: visual
---

## Overview

In this hands-on exercise, I learned how to plot functional and truthful choropleth maps by using the [**tmap**](https://r-tmap.github.io/tmap/) package. Choropleth mapping involves the symbolisation of enumeration units, such as countries, provinces, states, counties or census units, using area patterns or graduated colors.

## Getting Started

I loaded the required R packages: **sf** for handling geospatial data, **tmap** for thematic mapping, and **tidyverse** (which includes **readr**, **tidyr** and **dplyr**) for data import and wrangling.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Importing Data into R

### The Data

Two data sets were used to create the choropleth map:

-   Master Plan 2014 Subzone Boundary (Web) (i.e. `MP14_SUBZONE_WEB_PL`) in ESRI shapefile format. This is a geospatial data set consisting of the geographical boundary of Singapore at the planning subzone level, based on URA Master Plan 2014.

-   Singapore Residents by Planning Area / Subzone, Age Group, Sex and Type of Dwelling, June 2011-2020 in csv format (i.e. `respopagesextod2011to2020.csv`). This is an aspatial data file. Although it does not contain any coordinates values, its PA and SZ fields can be used as unique identifiers to geocode to `MP14_SUBZONE_WEB_PL` shapefile.

### Importing Geospatial Data into R

I used *st_read()* of **sf** package to import `MP14_SUBZONE_WEB_PL` shapefile into R as a simple feature data frame called `mpsz`.

```{r}
mpsz <- st_read(dsn = "data/geospatial",
                layer = "MP14_SUBZONE_WEB_PL")
```

I examined the content of `mpsz` below. Notice that only the first ten records are displayed.

```{r}
mpsz
```

### Importing Attribute Data into R

Next, I imported *respopagesextod2011to2020.csv* file into RStudio and saved it into an R dataframe called *popdata* using *read_csv()* of **readr** package.

```{r}
popdata <- read_csv("data/aspatial/respopagesextod2011to2020.csv")
```

### Data Preparation

Before preparing a thematic map, I needed to prepare a data table with year 2020 values. The data table should include the variables PA, SZ, YOUNG, ECONOMY ACTIVE, AGED, TOTAL, DEPENDENCY.

-   YOUNG: age group 0 to 4 until age group 20 to 24,
-   ECONOMY ACTIVE: age group 25-29 until age group 60-64,
-   AGED: age group 65 and above,
-   TOTAL: all age group, and
-   DEPENDENCY: the ratio between young and aged against economy active group

#### Data wrangling

I used the following data wrangling and transformation functions: *pivot_wider()* of **tidyr** package, and *mutate()*, *filter()*, *group_by()* and *select()* of **dplyr** package.

```{r}
popdata2020 <- popdata %>%
  filter(Time == 2020) %>%
  group_by(PA, SZ, AG) %>%
  summarise(`POP` = sum(`Pop`)) %>%
  ungroup() %>%
  pivot_wider(names_from=AG,
              values_from=POP) %>%
  mutate(YOUNG = rowSums(.[3:6])
         +rowSums(.[12])) %>%
mutate(`ECONOMY ACTIVE` = rowSums(.[7:11])+
rowSums(.[13:15]))%>%
mutate(`AGED`=rowSums(.[16:21])) %>%
mutate(`TOTAL`=rowSums(.[3:21])) %>%
mutate(`DEPENDENCY` = (`YOUNG` + `AGED`)
/`ECONOMY ACTIVE`) %>%
  select(`PA`, `SZ`, `YOUNG`,
       `ECONOMY ACTIVE`, `AGED`,
       `TOTAL`, `DEPENDENCY`)
```

#### Joining the attribute data and geospatial data

Before performing the georelational join, I converted the values in PA and SZ fields to uppercase, because the values of *PA* and *SZ* fields are made up of upper- and lowercase, while *SUBZONE_N* and *PLN_AREA_N* are in uppercase.

```{r}
popdata2020 <- popdata2020 %>%
  mutate(across(c(PA, SZ), toupper)) %>%
  filter(`ECONOMY ACTIVE` > 0)
```

Next, I used *left_join()* of **dplyr** to join the geographical data and attribute table using planning subzone name e.g. *SUBZONE_N* and *SZ* as the common identifier.

```{r}
mpsz_pop2020 <- left_join(mpsz, popdata2020,
                          by = c("SUBZONE_N" = "SZ"))
```

::: callout-note
*left_join()* of **dplyr** package is used with `mpsz` simple feature data frame as the left data table to ensure that the output will be a simple features data frame.
:::

```{r}
write_rds(mpsz_pop2020, "data/rds/mpszpop2020.rds")
```

## Choropleth Mapping Geospatial Data Using *tmap*

Two approaches can be used to prepare thematic maps using *tmap*:

-   Plotting a thematic map quickly by using *qtm()*.
-   Plotting highly customisable thematic map by using tmap elements.

### Plotting a choropleth map quickly by using *qtm()*

The easiest and quickest way to draw a choropleth map using **tmap** is *qtm()*. It is concise and provides a good default visualisation in many cases.

```{r}
tmap_mode("plot")
qtm(mpsz_pop2020,
    fill = "DEPENDENCY")
```

::: callout-note
-   *tmap_mode()* with "plot" option is used to produce a static map. For interactive mode, "view" option should be used.
-   *fill* argument is used to map the attribute (i.e. DEPENDENCY).
:::

### Creating a choropleth map by using *tmap*'s elements

Despite the usefulness of *qtm()* for drawing a choropleth map quickly, its disadvantage is that it makes aesthetics of individual layers harder to control. To draw a high quality cartographic choropleth map, **tmap**'s drawing elements should be used.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

In the following sub-sections, I will go through the tmap functions used to plot these elements.

#### Drawing a base map

The basic building block of **tmap** is *tm_shape()* followed by one or more layer elements such as *tm_fill()* and *tm_polygons()*. Here, *tm_shape()* defines the input data (i.e *mpsz_pop2020*) and *tm_polygons()* draws the planning subzone polygons.

```{r}
tm_shape(mpsz_pop2020) +
  tm_polygons()
```

#### Drawing a choropleth map using *tm_polygons()*

To draw a choropleth map showing the geographical distribution of a selected variable by planning subzone, I assigned the target variable *DEPENDENCY* to *tm_polygons()*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY")
```

::: callout-note
-   The default interval binning used is called "pretty".
-   The default colour scheme used is `YlOrRd` of ColorBrewer.
-   By default, missing values are shaded in grey.
:::

#### Drawing a choropleth map using *tm_fill()* and *tm_border()*

*tm_polygons()* is a wrapper of *tm_fill()* and *tm_border()*. *tm_fill()* shades the polygons by using the default colour scheme and *tm_borders()* adds the borders of the shapefile onto the choropleth map.

The code chunk below draws a choropleth map by using *tm_fill()* alone.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill("DEPENDENCY")
```

To add the boundary of the planning subzones, I used *tm_borders()*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY") +
  tm_borders(lwd = 0.01,
             fill_alpha = 0.1)
```

::: callout-note
The *alpha* argument defines transparency between 0 (totally transparent) and 1 (not transparent). Other arguments for *tm_borders()*: *col* (border colour), *lwd* (border line width, default is 1), and *lty* (border line type, default is "solid").
:::

### Data classification methods of **tmap**

**tmap** provides ten data classification methods: *fixed*, *sd*, *equal*, *pretty* (default), *quantile*, *kmeans*, *hclust*, *bclust*, *fisher*, and *jenks*.

#### Plotting choropleth maps with built-in classification methods

The code chunk below shows a Jenks data classification with 5 classes.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

Below I used *equal* data classification method.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "equal",
        n = 5)) +
  tm_borders(fill_alpha = 0.5)
```

I noticed that the distribution of quantile data classification method is more evenly distributed than equal data classification method.

#### Plotting choropleth map with custom breaks

For all the built-in styles, the category breaks are computed internally. To override these defaults, the breakpoints can be set explicitly by means of the *breaks* argument. In **tmap**, the breaks include a minimum and maximum, so in order to end up with n categories, n+1 elements must be specified.

First, I computed the descriptive statistics of **DEPENDENCY** field.

```{r}
summary(mpsz_pop2020$DEPENDENCY)
```

With reference to the results above, I set break points at 0.60, 0.70, 0.80, and 0.90, plus a minimum of 0 and maximum of 1.00. The *breaks* vector is thus c(0, 0.60, 0.70, 0.80, 0.90, 1.00).

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
          breaks = c(0, 0.60, 0.70, 0.80, 0.90, 1.00)) +
  tm_borders(fill_alpha = 0.5)
```

### Colour Scheme

**tmap** supports colour ramps either defined by the user or a set of predefined colour ramps from the **RColorBrewer** package.

#### Using ColourBrewer palette

To change the colour, I assigned the preferred colour to *values* argument of *tm_scale_intervals()*.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

To reverse the colour shading, I added a "-" prefix.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "quantile",
        n = 5,
        values = "-brewer.greens")) +
  tm_borders(fill_alpha = 0.5)
```

### Map Layouts

Map layout refers to the combination of all map elements into a cohesive map, including the objects to be mapped, the title, the scale bar, the compass, margins and aspect ratios.

#### Map Legend

In **tmap**, several `tm_legend()` options are provided to change the placement, format and appearance of the legend.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons("DEPENDENCY",
      fill.scale = tm_scale_intervals(
        style = "jenks",
        n = 5,
        values = "brewer.greens"),
      fill.legend = tm_legend(
        title = "Dependency ratio")) +
  tm_borders(fill_alpha = 0.5) +
  tm_title("Distribution of Dependency Ratio by planning subzone \n(Jenks classification)")
```

#### Cartographic Furniture

**tmap** also provides arguments to draw other map furniture such as compass, scale bar and grid lines. Below I used `tm_compass()`, `tm_scale_bar()` and `tm_grid()` to add compass, scale bar and grid lines onto the choropleth map.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(fill = "DEPENDENCY",
              fill.scale = tm_scale_intervals(
                style = "quantile",
                n = 5,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                title = "Dependency ratio")) +
  tm_title("Distribution of Dependency Ratio by planning subzone") +
  tm_layout(frame = TRUE) +
  tm_borders(fill_alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_grid(alpha =0.2) +
  tm_credits("Source: Planning Sub-zone boundary from Urban Redevelopment Authorithy (URA)\n and Population data from Department of Statistics DOS",
             position = c("left", "bottom"))
```

To reset the default style, I used the code chunk below.

```{r}
tmap_style("white")
```

### Drawing Small Multiple Choropleth Maps

**Small multiple maps**, also referred to as **facet maps**, are composed of many maps arranged side-by-side, and sometimes stacked vertically. They enable the visualisation of how spatial relationships change with respect to another variable, such as time.

In **tmap**, small multiple maps can be plotted in three ways:

-   by assigning multiple values to at least one of the aesthetic arguments,
-   by defining a group-by variable in *tm_facets()*, and
-   by creating multiple stand-alone maps with *tmap_arrange()*.

#### By assigning multiple values to at least one of the aesthetic arguments

In this example, I created small multiple choropleth maps by defining ***ncols*** in **tm_fill()**.

```{r}
tm_shape(mpsz_pop2020)+
  tm_fill(c("YOUNG", "AGED"),
          style = "equal",
          palette = "Blues") +
  tm_layout(legend.position = c("right", "bottom")) +
  tm_borders(alpha = 0.5) +
  tmap_style("white")
```

In the next example, I created small multiple choropleth maps by assigning multiple values to aesthetic arguments.

```{r}
tm_shape(mpsz_pop2020)+
  tm_polygons(c("DEPENDENCY","AGED"),
          style = c("equal", "quantile"),
          palette = list("Blues","Greens")) +
  tm_layout(legend.position = c("right", "bottom"))
```

#### By defining a group-by variable in *tm_facets()*

In this example, I created multiple small choropleth maps using `tm_facets()`.

```{r}
tm_shape(mpsz_pop2020) +
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          thres.poly = 0) +
  tm_facets(by="REGION_N",
            free.coords=TRUE) +
  tm_layout(legend.show = FALSE,
            title.position = c("center", "center"),
            title.size = 20) +
  tm_borders(alpha = 0.5)
```

#### By creating multiple stand-alone maps with `tmap_arrange()`

In this example, I created multiple small choropleth maps by creating multiple stand-alone maps with `tmap_arrange()`.

```{r}
youngmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("YOUNG",
              style = "quantile",
              palette = "Blues")

agedmap <- tm_shape(mpsz_pop2020)+
  tm_polygons("AGED",
              style = "quantile",
              palette = "Blues")

tmap_arrange(youngmap, agedmap, asp=1, ncol=2)
```

### Mapping Spatial Object Meeting a Selection Criterion

Instead of creating small multiple choropleth maps, I can also use a selection function to map spatial objects meeting a selection criterion.

```{r}
tm_shape(mpsz_pop2020[mpsz_pop2020$REGION_N=="CENTRAL REGION", ])+
  tm_fill("DEPENDENCY",
          style = "quantile",
          palette = "Blues",
          legend.hist = TRUE,
          legend.is.portrait = TRUE,
          legend.hist.z = 0.1) +
  tm_layout(legend.outside = TRUE,
            legend.height = 0.45,
            legend.width = 5.0,
            legend.position = c("right", "bottom"),
            frame = FALSE) +
  tm_borders(alpha = 0.5)
```

## Reference

### All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)
-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

### Geospatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
-   [Simple Features for R: Standardized Support for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

### Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)
-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
