---
title: "Analytical Mapping"
author: "Guofang Ji"
date: "March 6, 2026"
date-modified: "`r Sys.Date()`"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
editor: visual
---

## Overview

In this hands-on exercise, I learned how to plot analytical maps using appropriate R methods. By the end of this exercise, I was able to:

-   Import geospatial data in rds format into R environment.
-   Create cartographic quality choropleth maps by using appropriate tmap functions.
-   Create rate maps, percentile maps, and box maps.

## Getting Started

### Installing and loading packages

I loaded the required R packages: **sf**, **tmap** and **tidyverse**.

```{r}
pacman::p_load(tmap, tidyverse, sf)
```

### Importing data

For this hands-on exercise, I used a prepared data set called *NGA_wp.rds*. The data set is a polygon feature data.frame providing information on water points of Nigeria at the LGA level.

```{r}
NGA_wp <- read_rds("data/NGA_wp.rds")
```

## Basic Choropleth Mapping

### Visualising distribution of non-functional water point

I plotted choropleth maps showing the distribution of functional water points and total water points by LGA.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
p1 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "wp_functional",
             fill.scale = tm_scale_intervals(
               style = "equal",
               n = 10,
               values = "brewer.blues"),
             fill.legend = tm_legend(
               position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of functional water point by LGAs")
```

```{r}
#| code-fold: true
#| code-summary: "Show the code"
p2 <- tm_shape(NGA_wp) +
  tm_polygons(fill = "total_wp",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Distribution of total  water point by LGAs")
```

```{r}
#| fig-width: 16
tmap_arrange(p2, p1, nrow = 1)
```

## Choropleth Map for Rates

It is important to map rates rather than counts of things, because water points are not equally distributed in space. If we do not account for how many water points are somewhere, we end up mapping total water point size rather than our topic of interest.

### Deriving Proportion of Functional Water Points and Non-Functional Water Points

I tabulated the proportion of functional water points and the proportion of non-functional water points in each LGA. I used `mutate()` from **dplyr** package to derive two fields, namely *pct_functional* and *pct_nonfunctional*.

```{r}
NGA_wp <- NGA_wp %>%
  mutate(pct_functional = wp_functional/total_wp) %>%
  mutate(pct_nonfunctional = wp_nonfunctional/total_wp)
```

### Plotting map of rate

I plotted a choropleth map showing the distribution of percentage functional water point by LGA.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#| fig-height: 10
tm_shape(NGA_wp) +
  tm_polygons("pct_functional",
              fill.scale = tm_scale_intervals(
                style = "equal",
                n = 10,
                values = "brewer.blues"),
              fill.legend = tm_legend(
                position = c("right", "bottom"))) +
  tm_borders(lwd = 0.1,
             fill_alpha = 1) +
  tm_title("Rate map of functional water point by LGAs")
```

## Extreme Value Maps

Extreme value maps are variations of common choropleth maps where the classification is designed to highlight extreme values at the lower and upper end of the scale, with the goal of identifying outliers. These maps were developed in the spirit of spatializing EDA, i.e., adding spatial features to commonly used approaches in non-spatial EDA (Anselin 1994).

### Percentile Map

The percentile map is a special type of quantile map with six specific categories: 0-1%, 1-10%, 10-50%, 50-90%, 90-99%, and 99-100%. The corresponding breakpoints can be derived by means of the base R quantile command, passing an explicit vector of cumulative probabilities as c(0,.01,.1,.5,.9,.99,1).

#### Data Preparation

Step 1: I excluded records with NA.

```{r}
NGA_wp <- NGA_wp %>%
  drop_na()
```

Step 2: I created customised classification and extracted values.

```{r}
percent <- c(0,.01,.1,.5,.9,.99,1)
var <- NGA_wp["pct_functional"] %>%
  st_set_geometry(NULL)
quantile(var[,1], percent)
```

::: callout-note
When variables are extracted from an sf data.frame, the geometry is extracted as well. Many base R functions cannot deal with the geometry. Specifically, `quantile()` gives an error. As a result `st_set_geometry(NULL)` is used to drop the geometry field.
:::

#### Writing functions

Writing a function has three big advantages over using copy-and-paste:

-   You can give a function an evocative name that makes your code easier to understand.
-   As requirements change, you only need to update code in one place, instead of many.
-   You eliminate the chance of making incidental mistakes when you copy and paste.

#### Creating the get.var function

I wrote an R function to extract a variable (i.e. *wp_nonfunctional*) as a vector out of an sf data.frame.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>%
    st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### A percentile mapping function

Next, I wrote a percentile mapping function.

```{r}
percentmap <- function(vnam, df, legtitle=NA, mtitle="Percentile Map"){
  percent <- c(0,.01,.1,.5,.9,.99,1)
  var <- get.var(vnam, df)
  bperc <- quantile(var, percent)
  tm_shape(df) +
  tm_polygons() +
  tm_shape(df) +
     tm_polygons(vnam,
             title=legtitle,
             breaks=bperc,
             palette="Blues",
          labels=c("< 1%", "1% - 10%", "10% - 50%", "50% - 90%", "90% - 99%", "> 99%"))  +
  tm_borders() +
  tm_layout(main.title = mtitle,
            title.position = c("right","bottom"))
}
```

#### Test drive the percentile mapping function

I tested the function with the "total_wp" variable.

```{r}
percentmap("total_wp", NGA_wp)
```

### Box map

A box map is an augmented quartile map, with an additional lower and upper category. When there are lower outliers, the starting point for the breaks is the minimum value, and the second break is the lower fence. When there are no lower outliers, the starting point is the lower fence, and the second break is the minimum value.

```{r}
#| fig-width: 4
ggplot(data = NGA_wp,
       aes(x = "",
           y = wp_nonfunctional)) +
  geom_boxplot()
```

#### Creating the boxbreaks function

I wrote an R function to create break points for a box map.

```{r}
boxbreaks <- function(v,mult=1.5) {
  qv <- unname(quantile(v))
  iqr <- qv[4] - qv[2]
  upfence <- qv[4] + mult * iqr
  lofence <- qv[2] - mult * iqr
  # initialize break points vector
  bb <- vector(mode="numeric",length=7)
  # logic for lower and upper fences
  if (lofence < qv[1]) {  # no lower outliers
    bb[1] <- lofence
    bb[2] <- floor(qv[1])
  } else {
    bb[2] <- lofence
    bb[1] <- qv[1]
  }
  if (upfence > qv[5]) { # no upper outliers
    bb[7] <- upfence
    bb[6] <- ceiling(qv[5])
  } else {
    bb[6] <- upfence
    bb[7] <- qv[5]
  }
  bb[3:5] <- qv[2:4]
  return(bb)
}
```

#### Creating the get.var function

I wrote another version of the get.var function to extract a variable as a vector out of an sf data frame.

```{r}
get.var <- function(vname,df) {
  v <- df[vname] %>% st_set_geometry(NULL)
  v <- unname(v[,1])
  return(v)
}
```

#### Test drive the newly created function

I tested the newly created function.

```{r}
var <- get.var("wp_nonfunctional", NGA_wp)
boxbreaks(var)
```

#### Boxmap function

I wrote an R function to create a box map.

```{r}
boxmap <- function(vnam, df,
                   legtitle=NA,
                   mtitle="Box Map",
                   mult=1.5){
  var <- get.var(vnam,df)
  bb <- boxbreaks(var)
  tm_shape(df) +
    tm_polygons() +
  tm_shape(df) +
     tm_fill(vnam,title=legtitle,
             breaks=bb,
             palette="Blues",
          labels = c("lower outlier",
                     "< 25%",
                     "25% - 50%",
                     "50% - 75%",
                     "> 75%",
                     "upper outlier"))  +
  tm_borders() +
  tm_layout(main.title = mtitle,
            title.position = c("left",
                               "top"))
}
```

Finally, I tested the boxmap function.

```{r}
tmap_mode("plot")
boxmap("wp_nonfunctional", NGA_wp)
```
