---
title: "Visualising Geospatial Point Data"
author: "Guofang Ji"
date: "March 6, 2026"
date-modified: "`r Sys.Date()`"
format: html
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
editor: visual
---

## Overview

In this hands-on exercise, I learned how to create proportional symbol maps (also known as graduated symbol maps) showing the number of wins by Singapore Pools' outlets using the **tmap** package. Proportional symbol maps use the visual variable of size to represent differences in the magnitude of a discrete, abruptly changing phenomenon, e.g. counts of people.

By the end of this hands-on exercise, I acquired the following skills:

-   Importing an aspatial data file into R.
-   Converting it into a simple point feature data frame and assigning an appropriate projection reference.
-   Plotting interactive proportional symbol maps.

## Getting Started

I loaded the required R packages: **sf**, **tmap** and **tidyverse**.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

## Geospatial Data Wrangling

### The data

The data set used for this hands-on exercise is called *SGPools_svy21*. The data is in csv file format. It consists of seven columns. The XCOORD and YCOORD columns are the x-coordinates and y-coordinates of SingPools outlets and branches in [Singapore SVY21 Projected Coordinates System](https://www.sla.gov.sg/sirent/CoordinateSystems.aspx).

### Data Import and Preparation

I used *read_csv()* function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called *sgpools*.

```{r}
sgpools <- read_csv("data/SGPools_svy21.csv")
```

After importing the data file, I examined the data using *list()*.

```{r}
list(sgpools)
```

Notice that the *sgpools* data is in tibble data frame format.

### Creating a **sf** data frame from an aspatial data frame

I converted the sgpools data frame into a simple feature data frame by using *st_as_sf()* of **sf** package.

```{r}
sgpools_sf <- st_as_sf(sgpools,
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

::: callout-note
-   The *coords* argument requires the column name of the x-coordinates first, followed by the column name of the y-coordinates.
-   The *crs* argument requires the coordinates system in epsg format. [EPSG: 3414](https://epsg.io/3414) is Singapore SVY21 Projected Coordinate System.
:::

I displayed the basic information of the newly created sgpools_sf below. Notice that a new column called geometry has been added into the data frame, and the epsg ID is 3414.

```{r}
list(sgpools_sf)
```

## Drawing Proportional Symbol Map

To create an interactive proportional symbol map in R, I used the view mode of tmap.

```{r}
tmap_mode("view")
```

### An interactive point symbol map

I created an interactive point symbol map using the code chunk below.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
           size = 1,
           col = "black",
           lwd = 1)
```

### Making it proportional

To draw a proportional symbol map, I assigned the numerical variable *Gp1Gp2 Winnings* to the size visual attribute.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "red",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

### Giving it a different colour

I further improved the proportional symbol map by using the colour visual attribute. Here, the *OUTLET TYPE* variable is used as the colour attribute variable.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1)
```

### Faceted proportional symbol maps

An impressive feature of **tmap**'s view mode is that it also works with faceted plots. The argument *sync* in *tm_facets()* can be used to produce multiple maps with synchronised zoom and pan settings.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(fill = "OUTLET TYPE",
             size = "Gp1Gp2 Winnings",
             col = "black",
             lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```

Before ending the session, I switched **tmap**'s Viewer back to plot mode.

```{r}
tmap_mode("plot")
```

## Reference

### All about **tmap** package

-   [tmap: Thematic Maps in R](https://www.jstatsoft.org/article/view/v084i06)
-   [tmap](https://cran.r-project.org/web/packages/tmap/index.html)
-   [tmap: get started!](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
-   [tmap: changes in version 2.0](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-changes-v2.html)

### Geospatial data wrangling

-   [sf: Simple Features for R](https://cran.r-project.org/web/packages/sf/index.html)
-   [Simple Features for R: Standardized Support for Spatial Vector Data](https://journal.r-project.org/archive/2018/RJ-2018-009/RJ-2018-009.pdf)

### Data wrangling

-   [dplyr](https://dplyr.tidyverse.org/)
-   [Tidy data](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html)
