---
title: "Chapter 27: Modelling, Visualising and Analysing Network Data with R"
author: "Guofang Ji"
date: "March 10, 2026"
date-modified: "`r Sys.Date()`"
format:
  html:
    code-fold: true
    code-summary: "Show the code"
execute:
  eval: true
  echo: true
  warning: false
  freeze: true
editor: visual
---

## Overview

In this hands-on exercise, I learned how to model, analyse and visualise network data using R.

By the end of this hands-on exercise, I was able to:

-   create graph object data frames, manipulate them using appropriate functions of *dplyr*, *lubridate*, and *tidygraph*,
-   build network graph visualisation using appropriate functions of *ggraph*,
-   compute network geometrics using *tidygraph*,
-   build advanced graph visualisation by incorporating the network geometrics, and
-   build interactive network visualisation using *visNetwork* package.

## Getting Started

### Installing and launching R packages

I loaded four network data modelling and visualisation packages: **igraph**, **tidygraph**, **ggraph** and **visNetwork**. Beside these four packages, **tidyverse** and [**lubridate**](https://lubridate.tidyverse.org/), an R package specially designed to handle and wrangle time data, were also loaded. Additionally, **graphlayouts**, **concaveman** and **ggforce** were included to support advanced graph visualisation.

```{r}
pacman::p_load(igraph, tidygraph, ggraph,
               visNetwork, lubridate, clock,
               tidyverse, graphlayouts,
               concaveman, ggforce)
```

## The Data

The data sets used in this hands-on exercise are from an oil exploration and extraction company. There are two data sets. One contains the nodes data and the other contains the edges (also known as link) data.

### The edges data

*GAStech-email_edges.csv* consists of two weeks of 9063 email correspondences between 55 employees.

### The nodes data

*GAStech_email_nodes.csv* consists of the names, department and title of the 55 employees.

### Importing network data from files

I imported `GAStech_email_node.csv` and `GAStech_email_edges-v2.csv` into the RStudio environment using `read_csv()` of the **readr** package.

```{r}
GAStech_nodes <- read_csv("data/GAStech_email_node.csv")
GAStech_edges <- read_csv("data/GAStech_email_edge-v2.csv")
```

### Reviewing the imported data

I examined the structure of the data frame using *glimpse()* of **dplyr**.

```{r}
glimpse(GAStech_edges)
```

::: callout-warning
The output report of GAStech_edges above reveals that the *SentDate* is treated as "Character" data type instead of *date* data type. This is an error! Before continuing, it is important to change the data type of *SentDate* field back to "Date" data type.
:::

### Wrangling time

I used the code chunk below to perform the date type conversion.

```{r}
GAStech_edges <- GAStech_edges %>%
  mutate(SendDate = dmy(SentDate)) %>%
  mutate(Weekday = wday(SentDate,
                        label = TRUE,
                        abbr = FALSE))
```

::: callout-tip
## Things learned from the code chunk above

-   Both *dmy()* and *wday()* are functions of the **lubridate** package. [lubridate](https://cran.r-project.org/web/packages/lubridate/vignettes/lubridate.html) is an R package that makes it easier to work with dates and times.
-   *dmy()* transforms the SentDate to Date data type.
-   *wday()* returns the day of the week as a decimal number or an ordered factor if label is TRUE. The argument `abbr = FALSE` keeps the day spelled in full, i.e. Monday. The function creates a new column *Weekday* in the data frame.
-   The values in the *Weekday* field are in ordinal scale.
:::

### Reviewing the revised date fields

The table below shows the data structure of the reformatted *GAStech_edges* data frame.

```{r}
#| echo: false
glimpse(GAStech_edges)
```

### Wrangling attributes

A close examination of *GAStech_edges* data frame reveals that it consists of individual e-mail flow records. This is not very useful for visualisation.

In view of this, I aggregated the individual records by date, senders, receivers, main subject and day of the week.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  filter(MainSubject == "Work related") %>%
  group_by(source, target, Weekday) %>%
    summarise(Weight = n()) %>%
  filter(source!=target) %>%
  filter(Weight > 1) %>%
  ungroup()
```

::: callout-tip
## Things learned from the code chunk above

-   Four functions from **dplyr** package are used: *filter()*, *group_by()*, *summarise()*, and *ungroup()*.
-   The output data frame is called **GAStech_edges_aggregated**.
-   A new field called *Weight* has been added in GAStech_edges_aggregated.
:::

### Reviewing the revised edges file

The table below shows the data structure of the reformatted *GAStech_edges_aggregated* data frame.

```{r}
#| echo: false
glimpse(GAStech_edges_aggregated)
```

## Creating network objects using **tidygraph**

In this section, I learned how to create a graph data model using the **tidygraph** package. It provides a tidy API for graph/network manipulation. While network data itself is not tidy, it can be envisioned as two tidy tables — one for node data and one for edge data. **tidygraph** provides a way to switch between the two tables and exposes dplyr verbs for manipulating them, as well as access to a wide range of graph algorithms.

### The **tbl_graph** object

Two functions of **tidygraph** package can be used to create network objects:

-   [`tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) creates a **tbl_graph** network object from nodes and edges data.

-   [`as_tbl_graph()`](https://tidygraph.data-imaginist.com/reference/tbl_graph.html) converts network data and objects to a **tbl_graph** network.

### The **dplyr** verbs in **tidygraph**

The *activate()* verb from **tidygraph** serves as a switch between tibbles for nodes and edges. All dplyr verbs applied to a **tbl_graph** object are applied to the active tibble.

### Using `tbl_graph()` to build a tidygraph data model

I used `tbl_graph()` of the **tidygraph** package to build a tidygraph network graph data frame.

```{r}
GAStech_graph <- tbl_graph(nodes = GAStech_nodes,
                           edges = GAStech_edges_aggregated,
                           directed = TRUE)
```

### Reviewing the output tidygraph's graph object

```{r}
GAStech_graph
```

The output above reveals that *GAStech_graph* is a `tbl_graph` object with 54 nodes and 4541 edges. It prints the first six rows of "Node Data" and the first three rows of "Edge Data". The Node Data is shown as **active** by default.

### Changing the active object

The nodes tibble data frame is activated by default, but I can change which tibble is active with the *activate()* function. For example, to rearrange rows in the edges tibble to list those with the highest "weight" first:

```{r}
#| eval: false
GAStech_graph %>%
  activate(edges) %>%
  arrange(desc(Weight))
```

## Plotting Static Network Graphs with **ggraph** package

[**ggraph**](https://ggraph.data-imaginist.com/index.html) is an extension of **ggplot2**, making it easier to carry over basic ggplot skills to the design of network graphs.

There are three main aspects to a **ggraph** network graph:

-   [nodes](https://cran.r-project.org/web/packages/ggraph/vignettes/Nodes.html),
-   [edges](https://cran.r-project.org/web/packages/ggraph/vignettes/Edges.html), and
-   [layouts](https://cran.r-project.org/web/packages/ggraph/vignettes/Layouts.html).

### Plotting a basic network graph

I used [*ggraph()*](https://ggraph.data-imaginist.com/reference/ggraph.html), [*geom_edge_link()*](https://ggraph.data-imaginist.com/reference/geom_edge_link.html) and [*geom_node_point()*](https://ggraph.data-imaginist.com/reference/geom_node_point.html) to plot a basic network graph using *GAStech_graph*.

```{r}
ggraph(GAStech_graph) +
  geom_edge_link() +
  geom_node_point()
```

::: callout-tip
## Things learned from the code chunk above

-   The basic plotting function is `ggraph()`, which takes the data to be used for the graph and the type of layout desired. Both arguments for `ggraph()` are built around *igraph*. Therefore, `ggraph()` can use either an *igraph* object or a *tbl_graph* object.
:::

### Changing the default network graph theme

I used [*theme_graph()*](https://ggraph.data-imaginist.com/reference/theme_graph.html) to remove the x and y axes.

```{r}
g <- ggraph(GAStech_graph) +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

::: callout-tip
## Things learned from the code chunk above

-   **ggraph** introduces a special ggplot theme that provides better defaults for network graphs than the normal ggplot defaults. `theme_graph()` removes axes, grids, and borders, and changes the font to Arial Narrow.
-   The ggraph theme can be set for a series of plots with `set_graph_style()` run before the plots, or by using `theme_graph()` in each individual plot.
:::

### Changing the coloring of the plot

`theme_graph()` also makes it easy to change the coloring of the plot.

```{r}
g <- ggraph(GAStech_graph) +
  geom_edge_link(aes(colour = 'grey50')) +
  geom_node_point(aes(colour = 'grey40'))

g + theme_graph(background = 'grey10',
                text_colour = 'white')
```

### Fruchterman and Reingold layout

I plotted the network graph using the Fruchterman and Reingold layout by specifying `layout = "fr"`.

```{r}
g <- ggraph(GAStech_graph,
            layout = "fr") +
  geom_edge_link(aes()) +
  geom_node_point(aes())

g + theme_graph()
```

### Modifying network nodes

I coloured each node by its respective department using the `colour` aesthetic in *geom_node_point()*.

```{r}
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes()) +
  geom_node_point(aes(colour = Department,
                      size = 3))

g + theme_graph()
```

### Modifying edges

I mapped the thickness of the edges to the *Weight* variable using the `width` aesthetic in *geom_edge_link()*.

```{r}
g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 3)

g + theme_graph()
```

::: callout-tip
## Things learned from the code chunk above

-   *geom_edge_link* draws edges as straight lines between the start and end nodes. The `width` argument maps line width proportionally to the Weight attribute, and `alpha` introduces opacity to the lines.
:::

## Creating facet graphs

Another very useful feature of **ggraph** is faceting. In visualising network data, this technique can reduce edge over-plotting by spreading nodes and edges out based on their attributes.

There are three functions in **ggraph** to implement faceting:

-   [*facet_nodes()*](https://ggraph.data-imaginist.com/reference/facet_nodes.html) — edges are only drawn in a panel if both terminal nodes are present,
-   [*facet_edges()*](https://ggraph.data-imaginist.com/reference/facet_edges.html) — nodes are always drawn in all panels, and
-   [*facet_graph()*](https://ggraph.data-imaginist.com/reference/facet_graph.html) — faceting on two variables simultaneously.

### Working with *facet_edges()*

I used [*facet_edges()*](https://ggraph.data-imaginist.com/reference/facet_edges.html) to split the network by weekday.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_edges(~Weekday)
```

I then used *theme()* to move the legend to the bottom of the plot.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2) +
  theme(legend.position = 'bottom')

g + facet_edges(~Weekday)
```

### A framed facet graph

I added a frame to each facet panel using `th_foreground()`.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_edges(~Weekday) +
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

### Working with *facet_nodes()*

I used [*facet_nodes()*](https://ggraph.data-imaginist.com/reference/facet_nodes.html) to split the network by department.

```{r}
set_graph_style()

g <- ggraph(GAStech_graph,
            layout = "nicely") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department),
                  size = 2)

g + facet_nodes(~Department)+
  th_foreground(foreground = "grey80",
                border = TRUE) +
  theme(legend.position = 'bottom')
```

## Network Metrics Analysis

### Computing centrality indices

Centrality measures are a collection of statistical indices used to describe the relative importance of actors in a network. There are four well-known centrality measures: degree, betweenness, closeness and eigenvector.

I used *centrality_betweenness()* of **tidygraph** together with *mutate()* of **dplyr** to compute betweenness centrality and map node size accordingly.

```{r}
g <- GAStech_graph %>%
  mutate(betweenness_centrality = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
            size=betweenness_centrality))
g + theme_graph()
```

### Visualising network metrics

From **ggraph v2.0** onward, tidygraph algorithms such as centrality measures can be accessed directly in ggraph calls, without needing to precompute and store them.

```{r}
g <- GAStech_graph %>%
  ggraph(layout = "fr") +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(colour = Department,
                      size = centrality_betweenness()))
g + theme_graph()
```

### Visualising Community

**tidygraph** inherits many community detection algorithms from **igraph**, including *Edge-betweenness*, *Leading eigenvector*, *Fast-greedy*, *Louvain*, *Walktrap*, *Label propagation*, *InfoMAP*, *Spinglass*, and *Optimal*.

I used *group_edge_betweenness()* below to detect communities and colour the nodes accordingly.

```{r}
g <- GAStech_graph %>%
  mutate(community = as.factor(
    group_edge_betweenness(
      weights = Weight,
      directed = TRUE))) %>%
  ggraph(layout = "fr") +
  geom_edge_link(
    aes(
      width=Weight),
    alpha=0.2) +
  scale_edge_width(
    range = c(0.1, 5)) +
  geom_node_point(
    aes(colour = community))

g + theme_graph()
```

To support more effective visual investigation, I revised the community network using [`geom_mark_hull()`](https://ggforce.data-imaginist.com/reference/geom_mark_hull.html) from the **ggforce** package to draw convex hulls around each community group.

```{r}
#| fig-width: 10
#| fig-height: 10
g <- GAStech_graph %>%
  activate(nodes) %>%
  mutate(community = as.factor(
    group_optimal(weights = Weight)),
         betweenness_measure = centrality_betweenness()) %>%
  ggraph(layout = "fr") +
  geom_mark_hull(
    aes(x, y,
        group = community,
        fill = community),
    alpha = 0.2,
    expand = unit(0.3, "cm"),
    radius = unit(0.3, "cm")
  ) +
  geom_edge_link(aes(width=Weight),
                 alpha=0.2) +
  scale_edge_width(range = c(0.1, 5)) +
  geom_node_point(aes(fill = Department,
                      size = betweenness_measure),
                      color = "black",
                      shape = 21)

g + theme_graph()
```

## Building Interactive Network Graph with visNetwork

[visNetwork()](http://datastorm-open.github.io/visNetwork/) is an R package for network visualization using the [vis.js](http://visjs.org) javascript library.

*visNetwork()* uses a nodes list and edges list to create an interactive graph:

-   The nodes list must include an "id" column, and the edge list must have "from" and "to" columns.
-   The function also plots labels for the nodes using the "label" column in the node list.

The resulting graph allows interactive exploration — nodes can be moved, and the graph can be zoomed and panned.

### Data preparation

Before plotting the interactive network graph, I prepared the data model by joining the edges data with the nodes data to obtain numeric `from` and `to` IDs.

```{r}
GAStech_edges_aggregated <- GAStech_edges %>%
  left_join(GAStech_nodes, by = c("sourceLabel" = "label")) %>%
  rename(from = id) %>%
  left_join(GAStech_nodes, by = c("targetLabel" = "label")) %>%
  rename(to = id) %>%
  filter(MainSubject == "Work related") %>%
  group_by(from, to) %>%
    summarise(weight = n()) %>%
  filter(from!=to) %>%
  filter(weight > 1) %>%
  ungroup()
```

### Plotting the first interactive network graph

```{r eval=FALSE}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated)
```

### Working with layout

I applied the Fruchterman and Reingold layout using *visIgraphLayout()*.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr")
```

### Working with visual attributes - Nodes

*visNetwork()* looks for a field called "group" in the nodes object and colours the nodes according to the values of that field. I renamed the *Department* field to *group*.

```{r}
GAStech_nodes <- GAStech_nodes %>%
  rename(group = Department)
```

When I reran the code below, visNetwork shaded the nodes by assigning a unique colour to each category in the *group* field.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### Working with visual attributes - Edges

I used *visEdges()* to symbolise the edges with directional arrows and smooth curves.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visEdges(arrows = "to",
           smooth = list(enabled = TRUE,
                         type = "curvedCW")) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```

### Interactivity

I used *visOptions()* to add interactive features to the visualisation:

-   `highlightNearest = TRUE` highlights neighbouring nodes when a node is clicked.
-   `nodesIdSelection = TRUE` adds a dropdown selector for individual nodes.

```{r}
visNetwork(GAStech_nodes,
           GAStech_edges_aggregated) %>%
  visIgraphLayout(layout = "layout_with_fr") %>%
  visOptions(highlightNearest = TRUE,
             nodesIdSelection = TRUE) %>%
  visLegend() %>%
  visLayout(randomSeed = 123)
```
